Using an SMT engine to generate Symbolic Automata

### 2019-03-04 2017-Xudong-SMT/AVOCS-2018-postproc.tex
version Post Proceedings, soumise.

\documentclass{lncs/llncs}
\usepackage{etex}

\usepackage[T1]{fontenc}
% \usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
%\geometry{a4paper}                   % ... or a4paper or a5paper or ...
%\geometry{landscape}                % Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}

%\usepackage{amsfonts}
%\usepackage{fancyhdr}
%\usepackage{cite}
%\usepackage{ifthen}
%\usepackage{amssymb}
%\usepackage{fancyhdr}
%\usepackage{pifont}
\usepackage{stmaryrd}
\usepackage{mathtools,mathpartir}
\usepackage{proof}
%\usepackage{setspace}
\usepackage{enumitem}
\setlist{nolistsep}
%\usepackage{indentfirst}
\usepackage{amsmath,amssymb,amscd,mathrsfs}
% \usepackage{array,booktabs,arydshln,xcolor}
\usepackage{array,booktabs,xcolor}
\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}
\usepackage{epsfig,color,enumitem}
\usepackage[%
    font={small,sf},
    labelfont=bf,
    format=hang,    
    format=plain,
    margin=0pt,
    %% width=0.8\textwidth,
    aboveskip=4pt,
    belowskip=-18pt,
]{caption}
\usepackage[list=true]{subcaption}

@@

\usepackage{hyperref}
\usepackage{macrospNets}

@@

\newcommand{\sm}[1]{\mbox{\boldmath\small #1}}
\usepackage{algorithm} 
%\usepackage{algorithmicx} 
\usepackage{algpseudocode}  
\renewcommand{\algorithmicrequire}{\textbf{Input:}} 
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\algnewcommand{\IfThen}[2]{\State \algorithmicif\ #1\ \algorithmicthen\ #2}% \IfThenElse{<if>}{<then>}
\algnewcommand{\IfThenElse}[3]{\State \algorithmicif\ #1\ \algorithmicthen\ #2\ \algorithmicelse\ #3}% \IfThenElse{<if>}{<then>}{<else>}
%\DeclareMathOperator{\card}{card}
%\DeclareMathOperator{\Flat}{Flat}
\usepackage{listings}
\lstset{basicstyle=\small}
%\usepackage{algorithm} 
%\usepackage{algpseudocode}

\title{Using SMT engine to generate Symbolic Automata\thanks{This work was partially 
funded by the Associated Team FM4CPS
  between INRIA and ECNU, Shanghai}}
\author{ Xudong Qin\inst{1,2}  \ \ \  Simon Bliudze\inst{3} \ \ \ Eric Madelaine\inst{1}
  \ \ \  Min Zhang\inst{2}} 
\institute{Universit\'e C\^ote d'Azur, Inria, CNRS, I3S, 06902 Sophia Antipolis, France
        \and Shanghai Key Laboratory of Trustworthy Computing, ECNU, China
\and INRIA Lille -- Nord Europe, 40 avenue Halley,
59650 Villeneuve d'Ascq,
France}
\date{}                                           % Activate to display a given date or no date

@@

@@

Parameterized Networks of Synchronized Automata (pNets) were proposed
to give a behavioural specification formalism for distributed
systems, synchronous, asynchronous, or heterogeneous. They are used in
VerCors~\cite{HKLM:Foclasa14}, a platform for designing and 
verifying distributed systems, as the intermediate language for various
high-level languages. The high-level languages in VerCors formalize
each component of the distributed system and their composition.
pNets provides the core low-level semantic formalism for VerCors, and
is made of a hierarchical composition of (value-passing) automata,
called parameterized labelled transition systems (pLTS), where each
hierarchical level defines the possible synchronization of the lower levels.
Traditionally, pNets have been used to formalize fully
defined systems or softwares. But we want also to define and reason
about incompletely defined systems, like program skeletons, operators,
or open expressions of process calculi.
The open pNet model addresses this problem, using
"holes" as process parameters, representing \emph{unspecified} subsystems.
The pNet model was developed in a series of
papers~\cite{HMZ:PDP15,henrio:Forte2016} in which many examples have been
introduced showing its ability to encode the operators from some
other algebras or  program skeletons.
The operational semantics of an (open) pNet is defined as an
Open Automaton in which Open Transitions contain logical predicates
expressing the relations between the behaviour of the holes, and the
global behaviour of the system. In the previous publication,
only a sketch of a procedure allowing to compute these semantics was
presented, together with a proof of finiteness of the open automaton, under
reasonable hypotheses on the pNet structure.

@@

In the long term, we want to be able to check the equivalence between
open systems encoded as pNets. The equivalence between pNets is
"FH-bisimulation" \cite{henrio:Forte2016}, a dedicated version of
symbolic bisimulation taking the  
predicate of the open transitions into account when matching such
open transitions. We foresee that the interplay with the SMT solver
that we use here for satisfiability of open transitions will be
similar with what we need when proving (symbolic) equivalence between open
transitions.

@@

In \cite{henrio:Forte2016} we also
proved that a pNet with finitely many pLTSs, holes, and
synchronisation vectors, and with finite pLTSs, has a finite
semantics. Remark that all the elements of such pNets and open automata being
symbolic, they can represent many classes of unbounded systems.

\section{Generation of Open Automata}
\label{section:implementation}
In this section we describe the algorithm implementing the pNet
semantics, the interaction with the Z3 SMT solver, and we show the
result on our  example.

%% \noteSB{
%% line 11: Would it not be better to prune unsatisfiable transitions
%% on-the-fly, \ie inside the loop?}
%% \noteEM{Not so easy, it would require a proof... on one side this
%%   would implement reachability on the fly; but there is the question
%%   of collecting the assignments of the state variables, that are used
%%   in the sat check !}
Algorithm~\ref{alg1} starts with an open pNet, and builds its set of open
transitions. Its main loop is a classical residual algorithm: starting
from the initial global state, it picks a state in an unexplored set, and
computes all possible OTs, adding their target states in the
unexplored set, until this set is empty. As the set of
global states is finite (under the conditions above), this terminates.

@@

@@

Figure~\ref{schema:smt-lib} shows the translation of the
transition of Fig.~\ref{schema:unsat-ot}
in the SMTlib syntax.
It contains the declaration
of the BIP action algebra sorts 
and constructors, then the declaration of variables, and finally the
predicate to be checked, encoded as a set of assertions.
Here we display also the diagnosis (``sat'' or ``unsat'') generated by Z3;
%% \noteSB{I am not sure I understand the entire message of this phrase}\hl{though in practice our tool only needs the brute information ``sat''
%% or ``unsat''.
%% The result "unsat" in the output is just what we expected.
%% If the result is ``sat'', and more details are needed, Z3 can exhibit a
%% model (i.e. an instanciation of the variables) showing the satisfiability.
% as appears in the ``unsat'' proof generated by Z3 in which ``\{= ?x66 ?x70\}'' is shown to be false (see Fig. \ref{schema:smt-lib})

@@

@@

%\newpage
%\appendix
%\section{Appendix}
The following appendices are included for the convenience of reviewers.
Our intention is to publish as a research report, before final publication, an
extended version containing all technical details, proofs, and more examples.

@@

\begin{theorem}[Finiteness of open-automata.]\\
Given an open pNet $\mylangle \overline{\pNet},\overline{S}, \symb{SV}_k^{k\in K}\myrangle$ with leaves $pLTS_i^{i\in L}$ and holes $Hole_j^{j\in
  J}$, if the sets $L$ and $J$ are finite, if the synchronization vectors of all pNets 
  included in  $\mylangle \overline{\pNet},\overline{S}, \symb{SV}_k^{k\in K}\myrangle$ 
  are finite, and if
$\forall i \in L.\, finite{(states(pLTS_i))} \text{ and } pLTS_i$
has a finite number of state variables, then Algorithm 1 terminates
and produces an open automaton 
$\mathcal{T}$ with finitely many states and transitions.

>>>>

\documentclass{eceasst}
\usepackage{etex}

\usepackage[T1]{fontenc}
% \usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
%\geometry{a4paper}                   % ... or a4paper or a5paper or ...
%\geometry{landscape}                % Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent

%%\usepackage{graphicx}

%\usepackage{amsfonts}
%\usepackage{fancyhdr}
%\usepackage{cite}
%\usepackage{ifthen}
%\usepackage{amssymb}
%\usepackage{fancyhdr}
%\usepackage{pifont}
\usepackage{stmaryrd}
\usepackage{mathtools,mathpartir}
\usepackage{proof}
%\usepackage{setspace}
\usepackage{enumitem}
\setlist{nolistsep}
%\usepackage{indentfirst}
%%%\usepackage{amsmath,amssymb,amscd,mathrsfs}
% \usepackage{array,booktabs,arydshln,xcolor}
%
\usepackage{array,booktabs}
\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}
%\usepackage{epsfig,color,enumitem}
\usepackage{epsfig,enumitem}
\usepackage[%
    font={small,sf},
    labelfont=bf,
    format=hang,    
    format=plain,
    margin=0pt,
    %% width=0.8\textwidth,
    aboveskip=4pt,
    belowskip=-18pt,
]{caption}
\usepackage[list=true]{subcaption}

@@

%%\usepackage{hyperref}
\usepackage{macrospNets}

@@

\newcommand{\sm}[1]{\mbox{\boldmath\small #1}}
\usepackage{algorithm} 
%\usepackage{algorithmicx} 
\usepackage{algpseudocode}  
\renewcommand{\algorithmicrequire}{\textbf{Input:}} 
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\algnewcommand{\IfThen}[2]{\State \algorithmicif\ #1\ \algorithmicthen\ #2}% \IfThenElse{<if>}{<then>}
\algnewcommand{\IfThenElse}[3]{\State \algorithmicif\ #1\ \algorithmicthen\ #2\ \algorithmicelse\ #3}% \IfThenElse{<if>}{<then>}{<else>}
%\DeclareMathOperator{\card}{card}
%\DeclareMathOperator{\Flat}{Flat}
\usepackage{listings}

\lstset{basicstyle=\small}
%\usepackage{algorithm} 
%\usepackage{algpseudocode}

% Volume frontmatter
% ==================
\input{eceasst/frontmatter}

\title{Using an SMT engine to generate Symbolic Automata}
\author{ Xudong Qin\autref{1}\sponsor{This work was partially funded by the Associated Team FM4CPS between INRIA and ECNU, Shanghai},
  Simon Bliudze\autref{2}, Eric Madelaine\autref{3}, and Min Zhang\autref{4}} 
\institute{\autlabel{1} \email{marsxd@gmail.com}\\
  \autlabel{4} \email{mzhang@sei.ecnu.edu.cn}\\
  Shanghai Key Laboratory of Trustworthy Computing, ECNU, China\par
  \autlabel{2} \email{simon.bliudze@inria.fr}\\
  INRIA Lille -- Nord Europe, 40 avenue Halley, 59650 Villeneuve d'Ascq,
France\par
  \autlabel{3} \email{eric.madelaine@inria.fr}\\
  Universit\'e C\^ote d'Azur, Inria, CNRS, I3S, 06902 Sophia Antipolis, France
}
\date{}                                           % Activate to display a given date or no date

@@

\abstract{Open pNets are used to model the behaviour of open systems, both
synchronous or asynchronous, expressed in various calculi or
languages. They are endowed with a symbolic operational semantics in
terms of so-called ``Open Automata''. This allows us to check properties of
such systems in a compositional manner. 
We implement an algorithm computing these semantics,
building predicates expressing the synchronization conditions
between the events of the pNet sub-systems. Checking such
predicates requires symbolic reasoning over first order logics, but
also over application-specific data. We use the Z3 SMT engine to check
satisfiability of the predicates, and prune the open automaton of its
unsatisfiable transitions.
As an industrial oriented use-case, we use so-called ``architectures''
for BIP systems, that have been used in the framework of an ESA project and to specify the control software of
a nanosatellite at the EPFL Space Engineering Center. We use pNets to encode a BIP architecture
extended with explicit data, and compute its
open automaton semantics. This automaton may be used to prove
behavioural properties; we give 2 examples, a safety and a liveness
property.}
\keywords{Networks of Synchronised Automata, Symbolic Behavioral Semantics, Compositional Analysis of Software Components, SMT}

@@

Parameterized Networks of Synchronized Automata (pNets) were proposed
to give a behavioural specification formalism for distributed
systems, synchronous, asynchronous, or heterogeneous. They are used in
VerCors~\cite{HKLM:Foclasa14}, a platform for designing and 
verifying distributed systems, as the intermediate language for various
high-level languages. The high-level languages in VerCors formalize
each component of the distributed system and their composition.
pNets provides the core low-level semantic formalism for VerCors, and
is made of a hierarchical composition of (value-passing) automata,
called parameterized labelled transition systems (pLTS), where each
hierarchical level defines the possible synchronization of the lower levels.
Traditionally, pNets have been used to formalize fully
defined systems or softwares. But we want also to define and reason
about incompletely defined systems, like program skeletons, operators,
or open expressions of process calculi.
The open pNet model addresses this problem, using
``holes'' as process parameters, representing \emph{unspecified} subsystems.
The pNet model was developed in a series of
papers~\cite{HMZ:PDP15,henrio:Forte2016} in which many examples have been
introduced showing its ability to encode the operators from some
other algebras or  program skeletons.
The operational semantics of an (open) pNet is defined as an
Open Automaton in which Open Transitions contain logical predicates
expressing the relations between the behaviour of the holes, and the
global behaviour of the system. In the previous publication,
only a sketch of a procedure allowing to compute these semantics was
presented, together with a proof of finiteness of the open automaton, under
reasonable hypotheses on the pNet structure.

@@

In the long term, we want to be able to check the equivalence between
open systems encoded as pNets. The equivalence between pNets is
``FH-bisimulation'' \cite{henrio:Forte2016}, a dedicated version of
symbolic bisimulation taking the  
predicate of the open transitions into account when matching such
open transitions. We foresee that the interplay with the SMT solver
that we use here for satisfiability of open transitions will be
similar with what we need when proving (symbolic) equivalence between open
transitions.

@@

In \cite{henrio:Forte2016} we also proved:
\begin{theorem}[Finiteness]
  \label{Th:finiteness}
  Given an open pNet $pnet=\mylangle \overline{\pNet},\overline{S}, \symb{SV}_k^{k\in K}\myrangle$ with leaves $l_i^{i\in I}$ and holes $h_j^{j\in
  J}$, if the sets $I$ and $J$ are finite, if the synchronisation vectors of all pNets 
  included in  $pnet$   are finite, and if
$\forall i \in L.\, finite{(states(l_i))} \text{ and } l_i$
has a finite number of state variables, then its semantics is an open automaton 
$\mathcal{T}$ with finitely many states and transitions.
\end{theorem}
  Remark that all the elements of such pNets and open automata being
symbolic, they can represent many classes of unbounded systems.

\section{Generation of Open Automata}
\label{section:implementation}
In this section we describe the algorithm implementing the pNet
semantics, the interaction with the Z3 SMT solver, and we show the
result on our example. Under the hypotheses of Theorem \ref{Th:finiteness}
this algorithm terminates.

%% \noteSB{
%% line 11: Would it not be better to prune unsatisfiable transitions
%% on-the-fly, \ie inside the loop?}
%% \noteEM{Not so easy, it would require a proof... on one side this
%%   would implement reachability on the fly; but there is the question
%%   of collecting the assignments of the state variables, that are used
%%   in the sat check !}

@@

Algorithm~\ref{alg1} starts with an open pNet, and builds its set of open
transitions. Its main loop is a classical residual algorithm: starting
from the initial global state, it picks a state in an unexplored set, and
computes all possible OTs, adding their target states in the
unexplored set, until this set is empty.
%By construction, due to the
%satisfiability test in line 6, it only produces the set of states reachable
%through a sequence of satisfiable open transitions.

@@

Figure~\ref{schema:smt-lib} shows the translation of the
transition of Fig.~\ref{schema:unsat-ot}
in the SMTlib syntax.
It contains the declaration
of the BIP action algebra sorts 
and constructors, then the declaration of variables, and finally the
predicate to be checked, encoded as a set of assertions.
Here the behavior of the hole ``B'' appears as a variable
``|hb\_B:13:1|'', and line 10 shows the constraint that matching with
a synchronisation vector has imposed on the behavior of ``B''.
Here we display also the diagnosis (``sat'' or ``unsat'') generated by Z3.

@@

By lack of space, we have only shown here a use-case with boolean data
parameters, so we only require the basic SAT capability of the Z3
engine. It should be clear that more involved examples will require
reasonning on predicates of various data types. Z3 provides theories for reasonning
on integers, bitstrings, enumerated types, arrays, etc.
We also need to deal with some universal quantification on free
variables occuring in the guards of synch vectors.
For other data structures usual in such models, like records, or
recursive structures, the user will have to develop its own theories.
Decidability of these theories will condition the completeness of the
satisfiability check, and the decidability of model-checking or
bisimulation checking.

%% \noteSB{I am not sure I understand the entire message of this phrase}\hl{though in practice our tool only needs the brute information ``sat''
%% or ``unsat''.
%% The result "unsat" in the output is just what we expected.
%% If the result is ``sat'', and more details are needed, Z3 can exhibit a
%% model (i.e. an instanciation of the variables) showing the satisfiability.
% as appears in the ``unsat'' proof generated by Z3 in which ``\{= ?x66 ?x70\}'' is shown to be false (see Fig. \ref{schema:smt-lib})

@@

%\newpage
\appendix
\section{Changes from pre-proceedings version}
All minor requests from reviewers (typo, grammar) were already corrected in the preproceedings version.

Request from Reviewer 3, that we had not done in the preproceedings, for matter of space, have now been shortly answered.
More precisely,
\begin{itemize}
\item{} in section \ref{section:op-semantics}, we have added a precise statement of our finiteness theorem, that answers the question about termination of the algorithm.
\item{-} in section \ref{section:pruning} we have added more explanations about the way that various data types theories will be handled by the SMT engine
\end{itemize}

The following appendices are included for the convenience of reviewers.
Our intention is to publish as a research report, before final publication, an
extended version containing all technical details, proofs, and more examples.

@@

\begin{theorem}[Finiteness of open-automata.]\\
Given an open pNet $pnet=\mylangle \overline{\pNet},\overline{S}, \symb{SV}_k^{k\in K}\myrangle$ with leaves $l_i^{i\in L}$ and holes $h_j^{j\in
  J}$, if the sets $L$ and $J$ are finite, if the synchronization vectors of all pNets 
  included in  $pnet$ 
  are finite, and if
$\forall i \in L.\, finite{(states(l_i))} \text{ and } l_i$
has a finite number of state variables, then Algorithm 1 terminates
and produces an open automaton 
$\mathcal{T}$ with finitely many states and transitions.
